---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# stairtools

<!-- badges: start -->
<!-- badges: end -->

stairtools provides tools to compute and visualize straight stair geometries from basic architectural constraints.

The package separates the calculation into three main steps:

1. stair rise: determines the number of rises and the riser height from the total height.
2. stair run: computes the horizontal development of the flight.
3. stair geometry: builds the complete stepped profile.

## Design principle

The main function is `stair_compute()`, which combines these steps into a complete stair solution.

`max_run` represents the maximum available horizontal space, not a length that must be fully occupied. If this maximum theoretical value indicated does not fit in the available space, it is reduced accordingly. This means that a large available length does not automatically produce a longer stair: the unused space remains available.

The stair geometry follows the Blondel - comfort - relationship:

$$
2r + g = B
$$

where $r$ is the riser height, $g$ is the tread depth (going), and $B$ is the target Blondel value.


## Installation

Install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("remotes")

remotes::install_github("clement-LVD/stairtools")

``` 

```{r examples, eval = TRUE, echo = TRUE}

library(stairtools)

s <- stair_compute(height = 840, max_run = 1040)

plot(s$geometry)

```

The returned object contains:

- s$rise : vertical stair calculation
- s$build : horizontal stair calculation
- s$geometry : complete stair profile
- s$unused_run : remaining available horizontal space

The plot displays:

- the stepped profile,
- departure and arrival reference levels,
- cumulative horizontal dimensions,
- cumulative vertical dimensions.


```{r examples2}

s <- stair_compute(
  height = 3000,
  max_run = 6000,
  start = landing(depth = 1200),
  end = landing(depth = 1000)
)

str(s)

```

## Constraints

The computation accepts optional limits:

```{r examples3} 

s <- stair_compute(
  height = 3000,
  max_run = 6000,
  start = landing(depth = 1200),
  end = landing(depth = 1000)
)
 

```

Where:

- blondel_target controls the preferred comfort relationship,
- min_going prevents unrealistic very small treads,
- max_going limits excessively deep treads.

## Units

All dimensions are expressed in millimetres.