---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# stairtools

<!-- badges: start -->
[![R-CMD-check](https://github.com/Clement-LVD/stairtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Clement-LVD/stairtools/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Clement-LVD/stairtools/graph/badge.svg)](https://app.codecov.io/gh/Clement-LVD/stairtools)
<!-- badges: end -->

🪜 `stairtools` is an R package designed to calculate and visualise *staircase* designs, based on basic architectural constraints. Results are sorted according to a step
height optimization, in order to choose easily the best stair dimensioning.

## Installation

<!-- 
Install the released version from CRAN:
```{r eval = FALSE}
install.packages("stairtools")
```
-->

Install the latest development version from GitHub:

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("clement-LVD/stairtools")
```

## Preamble and usage

**㎝.** All dimensions are expressed in *centimetres*.

**Input.** The user must indicate a total rise to elevate and a total length available. If length is not a constraint on your project, you can specify a very large value.

**Output.** The package calculates all possible – reasonable – staircase configurations and return various solutions, i.e. varying numbers of steps and whether or not there is a landing step. 

> 𓊍 The possible solutions are scored according to Blondel's rule (see below).

**Blondel's value.** A stair geometry follows the Blondel - comfort - relationship.

$$
2r + g = B
$$

Where $r$ is the rise (vertical riser height), $g$ is the going (horizontal tread depth), and $B$ is the target Blondel value.

𓊍 In French carpentry and masonry practices for domestic staircase construction, the Blondel ideal value should be 63 cm. It is recommended to prioritise the staircase solution with the smallest deviation from this Blondel target value. From this French point of view, solutions with a Blondel value between 60 cm and 64 cm are acceptable. 

**Best solution.** Possible solutions are sorted by their deviation from the Blondel target value, default is 63 cm. 

**Edges cases.** When several solutions have a similar Blondel value, solutions are sorted by their deviation from the minimum rise, i.e. 16 cm. This is to ensure that the staircase is comfortable for older people, those wearing high-heeled shoes, children, small dogs and old dogs, etc. 

## Examples

The examples cover the various stair calculations provided by `solve_stairs()`, followed by the graphical parameters provided by `plot_stair()`.

### Stair computations

**Basic concrete stairs.** `solve_stairs()` compute various basic stair solutions, given a total height and a maximum horizontal run available. The function returns a `data.frame` listing several solutions, the most relevant of which is the first entry. A `list` variable called `geometry` provides the coordinates for each solution, with one entry for each solution in the main data.frame, in the same order.
```{r}

library(stairtools)

sol <- solve_stairs(total_height = 103, max_horizontal_run =  133)

print(sol)

# plot the best solution
plot_stair(sol$geometry[[1]]) 

```

Some staircase designs have a landing step at the top. For these solutions, thhe `has_landing` variable is set to `TRUE`.

```{r}

sol2 <- solve_stairs(total_height = 68, max_horizontal_run =  133)

landing_step_solutions <- sol2[sol2$has_landing == TRUE, ]
 
# plot the best solution within solutions providing a landing step
plot_stair(landing_step_solutions$geometry[[1]]) 

```

By default, `solve_stairs()` return basic concrete stairs, computed with `tread_thickness = 0, riser_thickness = 0, nosing = 0` and riser - vertical limit - are ploted with `riser = TRUE`, the default - as shown above. 

**Tread thickness and nosing.** The example below shows a 7 cm thick covering layer over a concrete staircase, extending outwards to leave a 4 cm stair nosing. The parameters are therefore `tread_thickness = 7, nosing = 4, riser_thickness = 0`.

```{r}

sol3 <- solve_stairs(80, 150, tread_thickness = 7, nosing = 4, riser_thickness = 0)

plot_stair(sol3$geometry[[1]], riser = TRUE)

```

When a stair nosing is required, the final top step is shorter by default - in order to maintain a constant going and thus avoid a ‘top-of-the-flight’ effect.

**Alternative nosing direction.** Nosing can instead be added in the negative direction with `positive_nosing_direction = FALSE`.
```{r}

sol_neg <- solve_stairs(80, 150, tread_thickness = 4, nosing = 2.5, positive_nosing_direction = FALSE)
 
plot_stair(sol_neg$geometry[[1]])

```

With this setting, all steps keep the same theoretical going, but the first step may protrude beyond the indicated available space. With the default positive direction, the last step is shortened instead, so that the staircase does not extend beyond the available run.

**Wood-stairs.** Wood stairs can be computed by specifying `tread_thickness` `riser_thickness` and or `nosing` parameters, expressed in centimeters.

```{r}

wood_stairs <- solve_stairs(80, 150, tread_thickness = 4, riser_thickness = 2.6, nosing = 2.5)

plot_stair(wood_stairs$geometry[[1]])

```

**Open-riser stairs.** Open-riser stairs can be plotted with `plot_stair(riser = FALSE)`. By default, `plot_stair()` displays the vertical risers.

```{r}
sol4 <- solve_stairs(80, 150, tread_thickness = 4,  nosing = 4)

plot_stair(sol4$geometry[[1]], riser = FALSE)
```

### Graphical parameters

`plot_stair()` uses base R's `graphics::polygon()` to draw each stair surface. Graphical parameters can therefore be passed through `polygon_params`.

For example, the colour and border of all stair surfaces can be changed.

```{r}

plot_stair(
 wood_stairs$geometry[[1]],
  polygon_params = list(
    col = "grey90",
    border = "black",
    lwd = 2
  )
)

```

Other graphical parameters supported by `graphics::polygon()` can also be used, including line types and hatching.

```{r}
plot_stair(
  wood_stairs$geometry[[1]],
  polygon_params = list(
    col = "black",
    border = "black",
    density = 25,
    angle = 45
  )
)
```

Styles can be applied selectively to specific steps or surfaces. The `surface` selector accepts "tread" or "riser", while steps can be a single step number or a vector of step numbers.

For example, treads and risers can be displayed differently.

```{r}
plot_stair(
  wood_stairs$geometry[[1]],
  styles = list(
    list(
      surface = "tread",
      col = "black",
      density = 25,
      angle = 45
    ),
    list(
      surface = "riser",
      col = "grey50"
    )
  )
)
```

Styles can also target individual step with `steps = c()`. 

```{r}
plot_stair(
  wood_stairs$geometry[[1]],
  styles = list( 
    list(
      steps = 1:4,
      col = "white",
      border = "black",
      lwd = 3
    ), list(
      steps = 1,
      surface = "riser", col = "black", density = 20,
    angle = 45 
    ),
    list(
      steps = 4,
      col = "black",
      border = "red",
      density = 15,
    angle = 45 
    )
  )
)

```

Styles are applied in order. When several styles apply to the same polygon, parameters defined by a later style replace parameters with the same name defined by earlier styles.

The polygon_params argument therefore provides a global default style if no `steps` is specified, while styles can be used to override individual steps or surfaces.

All graphical parameters accepted by graphics::polygon() can be passed in this way.

# Details
In the same vein, older French laws therefore stipulate that "the height and width must satisfy the relationship 0.60 m ≤ 2 H + G ≤ 0.64 m" (https://www.legifrance.gouv.fr/codes/article_lc/LEGIARTI000020272650/). Other standards and laws do not specify the permissible Blondel values, or even specify a range of values that differs from the range set out in French law, e.g., UK laws specify permissible Blondel values between 55 cm and 70 cm (https://assets.publishing.service.gov.uk/media/60d5bdcde90e07716f516cfd/Approved_Document_K.pdf). In other words, it is possible to build staircases that complying with British standards but do not comply with French standards.

Some standards apply specifically to a particular type of staircase, e.g., according to the ISO standard, a stair that is a permanent mean of access to machinery require a Blondel value between 60 cm and 66 cm^[International Organization for Standardization. (2016). Safety of machinery — Permanent means of access to machinery — Part 3: Stairs, stepladders and guard-rail (ISO Standard No. 14122-3:2016). https://www.iso.org/standard/61282.html].