---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# stairtools

<!-- badges: start -->
[![R-CMD-check](https://github.com/Clement-LVD/stairtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Clement-LVD/stairtools/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Clement-LVD/stairtools/graph/badge.svg)](https://app.codecov.io/gh/Clement-LVD/stairtools)
<!-- badges: end -->

🪜 `stairtools` is an R package designed to calculate and visualise *staircase* designs, based on basic architectural constraints. Results are sorted according to a step
height optimization, in order to choose easily the best stair dimensioning.

## Installation

<!-- 
Install the released version from CRAN:
```{r eval = FALSE}
install.packages("stairtools")
```
-->

Install the latest development version from GitHub:

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("clement-LVD/stairtools")
```

## Preamble and usage

**㎝.** All dimensions are expressed in *centimetres*.

**Input.** The user must indicate a total rise to elevate and a total length available. If length is not a constraint on your project, you can specify a very large value.

**Output.** The package calculates all possible – reasonable – staircase configurations and return various solutions, i.e. varying numbers of steps and whether or not there is a landing step. 

> 𓊍 The possible solutions are scored according to various rules and guidances (see below). 


## Examples

The examples cover the various stair calculations provided by `solve_stairs()`, followed by the graphical parameters provided by `plot_stair()`.

### Stair computations

**Basic concrete stairs.** `solve_stairs()` compute various basic stair solutions, given a total height and a maximum horizontal run available. The function returns a `data.frame` listing several solutions, the most relevant of which is the first entry. A `list` variable called `geometry` provides the coordinates for each solution, with one entry for each solution in the main data.frame, in the same order.
```{r}

library(stairtools)

sol <- solve_stairs(total_height = 103, max_horizontal_run =  133)

print(sol)

# plot the best solution
plot_stair(sol$geometry[[1]]) 

```

Some staircase designs have a landing step at the top. For these solutions, thhe `has_landing` variable is set to `TRUE`.

```{r}

sol2 <- solve_stairs(total_height = 68, max_horizontal_run =  133)

landing_step_solutions <- sol2[sol2$has_landing == TRUE, ]
 
# plot the best solution within solutions providing a landing step
plot_stair(landing_step_solutions$geometry[[1]]) 

```

By default, `solve_stairs()` return basic concrete stairs, computed with `tread_thickness = 0, riser_thickness = 0, nosing = 0` and riser - vertical limit - are ploted with `riser = TRUE`, the default - as shown above. 

**Tread thickness and nosing.** The example below shows a 7 cm thick covering layer over a concrete staircase, extending outwards to leave a 4 cm stair nosing. The parameters are therefore `tread_thickness = 7, nosing = 4, riser_thickness = 0`.

```{r}

sol3 <- solve_stairs(80, 150, tread_thickness = 7, nosing = 4, riser_thickness = 0)

plot_stair(sol3$geometry[[1]], riser = TRUE)

```

When a stair nosing is required, the final top step is shorter by default - in order to maintain a constant going and thus avoid a ‘top-of-the-flight’ effect.

**Alternative nosing direction.** Nosing can instead be added in the negative direction with `positive_nosing_direction = FALSE`.
```{r}

sol_neg <- solve_stairs(80, 150, tread_thickness = 4, nosing = 2.5, positive_nosing_direction = FALSE)
 
plot_stair(sol_neg$geometry[[1]])

```

With this setting, all the treads have the same dimensions, but the first step may protrude beyond the indicated available space. With the default positive direction, the last step is shortened instead, so that the staircase does not extend beyond the available run.

**Wood-stairs.** Wood stairs can be computed by specifying `tread_thickness` `riser_thickness` and or `nosing` parameters, expressed in centimeters.

```{r}

wood_stairs <- solve_stairs(80, 150, tread_thickness = 4, riser_thickness = 2.6, nosing = 2.5)

plot_stair(wood_stairs$geometry[[1]])

```

By default, the risers are positioned behind the tread, they do not overlap the nosing.

**Risers.** 

xxx todo : various risers computations xxx

**Open-riser stairs.** Open-riser stairs can be plotted with `plot_stair(riser = FALSE)`. By default, `plot_stair()` displays the vertical risers.

```{r}
sol4 <- solve_stairs(80, 150, tread_thickness = 4,  nosing = 4)

plot_stair(sol4$geometry[[1]], riser = FALSE)
```


# Legal and regulatory aspects

To qualify the solutions, the `check_stair_rules` parameters adds various logical columns relating to the legal or academic compliance of the proposed staircases.

```{r}

wood_stairs <- solve_stairs(180, 200, tread_thickness = 4, riser_thickness = 2.6, nosing = 2.5, check_stair_rules = TRUE)

# several logical columns are added, and 2 numeric (n_rules_ok & rate_rules_ok)
print(wood_stairs) 

plot_stair(wood_stairs$geometry[order(-wood_stairs$rate_rules_ok)][[1]] )


```

### Graphical parameters

`plot_stair()` uses base R's `graphics::polygon()` to draw each stair surface. Graphical parameters can therefore be passed through `polygon_params`.

For example, the colour and border of all stair surfaces can be changed.

```{r}

plot_stair(
 wood_stairs$geometry[[1]],
  polygon_params = list(
    col = "grey90",
    border = "black",
    lwd = 2
  )
)

```

Other graphical parameters supported by `graphics::polygon()` can also be used, including line types and hatching.

```{r}
plot_stair(
  wood_stairs$geometry[[1]],
  polygon_params = list(
    col = "black",
    border = "black",
    density = 25,
    angle = 45
  )
)
```

Styles can be applied selectively to specific steps or surfaces. The `surface` selector accepts "tread" or "riser", while steps can be a single step number or a vector of step numbers.

For example, treads and risers can be displayed differently.

```{r}
plot_stair(
  wood_stairs$geometry[[1]],
  styles = list(
    list(
      surface = "tread",
      col = "black",
      density = 25,
      angle = 45
    ),
    list(
      surface = "riser",
      col = "grey50"
    )
  )
)
```

Styles can also target individual step with `steps = c()`. 

```{r}
plot_stair(
  wood_stairs$geometry[[1]],
  styles = list( 
    list(
      steps = 1:4,
      col = "white",
      border = "black",
      lwd = 3
    ), list(
      steps = 1,
      surface = "riser", col = "black", density = 20,
    angle = 45 
    ),
    list(
      steps = 4,
      col = "black",
      border = "red",
      density = 15,
    angle = 45 
    )
  )
)

```

Styles are applied in order. When several styles apply to the same polygon, parameters defined by a later style replace parameters with the same name defined by earlier styles.

The polygon_params argument therefore provides a global default style if no `steps` is specified, while styles can be used to override individual steps or surfaces.

All graphical parameters accepted by graphics::polygon() can be passed in this way.

### Traditional stair computation

If no satisfactory solution is found, traditional stairs are computed. 

**Blondel solution.** Possible solutions are sorted by their deviation from the Blondel target value, default is 63 cm. When several solutions have a similar Blondel value, solutions are sorted by their deviation from the minimum rise (default is 16 cm). This is to ensure that the staircase is comfortable. 

**Blondel's value.** A stair geometry have a Blondel value.

$$
2r + g = B
$$

Where $r$ is the rise (vertical riser height), $g$ is the going (horizontal tread depth), and $B$ is the target Blondel value.

𓊍 In French carpentry and masonry traditional practices for domestic staircase construction, it is recommended to prioritise the staircase solution with the smallest deviation from a Blondel target value of 63 cm, and solutions with a Blondel value between 60 cm and 64 cm are acceptable. Therefore, it's important to note that this indicator is highly open to criticism because "reasoning behind the formula is dubious, and its applicability is demonstrably limited to a small number of average sizes, yet its acceptance by professionals and policy makers remains remarkably and unshakenly universal" (Koutamanis, 2024, p. 696). The Blondel value used as a reference by `solve_stairs` (63 cm) reflect the traditional French stair geometry criteria and does not allow compliance with most of the standards when risers heights > 17 cm are adopted. 
